---
title: "DATA 607 Project 2"
author: "Samuel I Kigamba"
date: "October 06, 2019"
output:
  html_document:
    code_folding: show
    df_print: paged
    highlight: pygments
    theme: united
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(stringr)
library(tidyr)
library(dplyr)
library(tidyverse)
library(tibble)
library(caret)
library(readr)
```
#--------------------------------------------------------------------------------

#\clearpage

DATA 607 Project 2.

The goal of this assignment is to give you practice in preparing different datasets for downstream
analysis work.
Your task is to:
(1) Choose any three of the "wide" datasets identified in the Week 5 Discussion items. (You may
use your own dataset; please don't use my Sample Post dataset, since that was used in your
Week 6 assignment!) For each of the three chosen datasets:
??? Create a .CSV file (or optionally, a MySQL database!) that includes all of the information
included in the dataset. You're encouraged to use a "wide" structure similar to how the
information appears in the discussion item, so that you can practice tidying and
transformations as described below.
??? Read the information from your .CSV file into R, and use tidyr and dplyr as needed to
tidy and transform your data. [Most of your grade will be based on this step!]
??? Perform the analysis requested in the discussion item.
??? Your code should be in an R Markdown file, posted to rpubs.com, and should include
narrative descriptions of your data cleanup work, analysis, and conclusions.
(2) Please include in your homework submission, for each of the three chosen datasets:
??? The URL to the .Rmd file in your GitHub repository, and
??? The URL for your rpubs.com web page.


set working directory and Install all the relevant packages and load their respective libraries into R.
```{r, include=FALSE}
getwd()
setwd("C:/Users/iguku/Google Drive/R and SQL/CUNY MSc Data Science/Data 607/Project 2")
```
#Migrants by destimation country

##Load the following libraries

####library(stringr)
####library(tidyr)
####library(dplyr)
####library(tidyverse)
####library(tibble)
####library(caret)
####library(readr)

##Upload the data into Github
This will ensure that everyone with access to the github repository can easily audit or retest the data. This ensures ease of accessibility and testing by a wide audience. Follow this link to see uploaded Male migrants .csv file (https://raw.githubusercontent.com/igukusamuel/DATA-607-Project-2/master/UN_MigrantStockBySexByDestination_2019.csv)

```{r message=FALSE, warning=FALSE}
migrants <- read_csv("https://raw.githubusercontent.com/igukusamuel/DATA-607-Project-2/master/UN_MigrantStockBySexByDestination_2019.csv")
head(migrants)

#view(head(male_migrants, 20)) # vIew data frame structure and see how many rows to skip.
```

##Skip first 15 rows
As part of data cleanup, skip the first 15 rows that include source information not relevant to out analysis.

```{r message=FALSE, warning=FALSE}
migrants <- read_csv("https://raw.githubusercontent.com/igukusamuel/DATA-607-Project-2/master/UN_MigrantStockBySexByDestination_2019.csv", skip = 15)

head(migrants) #Print out first few rows to confirm that the data have been loaded correctly.
```

##Filter for N/As in column X5
Careful review of the data shows that column named X5 only includes data for rows related to countries and N/A's for rows relating to regions and regional totals. Thus filtering out all N/As in column X5 will leave us with country data only, which is the basis of out analysis. We first view all the N/As under column X5 to confirm none of them relate to country information.

```{r message=FALSE, warning=FALSE}
colX5 <- filter(migrants, is.na(X5))

x <- length(colX5)
x

head(colX5)

```


##Exclude N/As in column X5
We then exclude all N/A's in column X6 and print out the first 6 rows using the head() function.

```{r message=FALSE, warning=FALSE}
migrants_by_country <- filter(migrants, !is.na(X5))

head(migrants_by_country)
```

##Rename column X2
From the above print out, there is need to rename column X2 dest_country.

```{r message=FALSE, warning=FALSE}
migrants_by_country <- migrants_by_country %>% 
        rename(
                dest_country = X2
        )
head(migrants_by_country)
```

##View all columns
The above printout shows a number of irrelevant columns that are not necessary for our analysis. Lets print out the entire column names and delete the unnecessary ones to have a cleaner data set.
```{r message=FALSE, warning=FALSE}
column_names <- colnames(migrants_by_country)
#column_names # umcomment to view entire list of column names
head(column_names)
```

##Exclude irrelevant columns
The above print out reveals that we do not need all column names that start with "X". We delete these columns using the srtarts_with function.

```{r message=FALSE, warning=FALSE}
migrants_by_country <- migrants_by_country %>% 
        select(-starts_with("X"))


migrants_by_country <- migrants_by_country %>% 
        select(-c(2:8))

migrants_by_country
```

##View dimentions of resulting data frame
We use dim() function to have an idea of how many rows and columns we have for our analysis.
```{r message=FALSE, warning=FALSE}
dim(migrants_by_country)
```

##Confrim column names.
This is what we need for our analysis.

```{r message=FALSE, warning=FALSE}
column_names_clean <- colnames(migrants_by_country)
#column_names_clean # uncomment to view entire list of cleaned up column names
head(column_names_clean)

```


##View number of columns
Get the length of the column names to be used in the next line of code.

```{r message=FALSE, warning=FALSE}
y <- length(colnames(migrants_by_country))

y 
```

##clean up data
Let us use gather() function to gather all columns with years into a single columns and exclude any and all N/As to obtain clean data. Spread the resulting data by year column and rename "1" as male and "2" as female.

```{r message=FALSE, warning=FALSE}
no_of_migrants_per_country <- mutate(gather(migrants_by_country, "year", "no_of_migrants", 2:y, na.rm = TRUE))

head(no_of_migrants_per_country)
```

```{r message=FALSE, warning=FALSE}
no_of_migrants_per_country <- no_of_migrants_per_country %>%
        separate(year, c("year", "sex"), sep = "_")

no_of_migrants_per_country
```

Convert the years column to number format

```{r message=FALSE, warning=FALSE}
no_of_migrants_per_country$year <- parse_number(no_of_migrants_per_country$year)

no_of_migrants_per_country
```

```{r message=FALSE, warning=FALSE}
no_of_migrants_per_country <- no_of_migrants_per_country %>%
        spread(sex, no_of_migrants)
        

names(no_of_migrants_per_country)
```

```{r message=FALSE, warning=FALSE}
no_of_migrants_per_country <- no_of_migrants_per_country %>% 
        rename(
                male = "1",
                female = "2"
        )
head(no_of_migrants_per_country)

```

##Conversion of chr to dbl
convert the no_of_migrants data column from characters to doubles for statistical analysis. This we will do using the parse_number() function. Print out using head() function the first 6 rows and confirm this conversion.

```{r message=FALSE, warning=FALSE}
no_of_migrants_per_country$male <- parse_number(no_of_migrants_per_country$male)
no_of_migrants_per_country$female <- parse_number(no_of_migrants_per_country$female)

clean_migrants_data <- no_of_migrants_per_country

head(clean_migrants_data)
```

#Down stream analysis

##Ordering of data
Ordering data by country with largest inflow of migrants

```{r message=FALSE, warning=FALSE}
by_country <- clean_migrants_data %>% 
        group_by(year, dest_country) %>% 
        summarise(total_migrants = male + female) %>% 
        arrange(desc(total_migrants))
head(by_country)
```

Ordering the data of male migrants by the destination countries by year

```{r message=FALSE, warning=FALSE}
male_by_country <- clean_migrants_data %>% 
        group_by(dest_country, year) %>% 
        summarise(male = male) %>% 
        arrange(desc(male))
head(male_by_country)
```

Ordering the data of female migrants by the destination countries by year
```{r message=FALSE, warning=FALSE}
female_by_country <- clean_migrants_data %>% 
        group_by(dest_country, year) %>% 
        summarise(female = female) %>% 
        arrange(desc(female))
head(female_by_country)
```

Ordering the data by % of male migrants by the destination countries by year

```{r message=FALSE, warning=FALSE}
Perc_male_by_country <- clean_migrants_data %>% 
        group_by(dest_country, year) %>% 
        summarise(perc_male = male/(male + female)) %>% 
        arrange(desc(perc_male))
head(Perc_male_by_country)
```

Ordering the data by % female migrants by the destination countries by year
```{r message=FALSE, warning=FALSE}
Perc_female_by_country <- clean_migrants_data %>% 
        group_by(dest_country, year) %>% 
        summarise(perc_female = female/(male + female)) %>% 
        arrange(desc(perc_female))
head(Perc_female_by_country)
```

#Conclusion
Maldives received the highest % of male migrants while nepal received the highest % of female migrants